Quasi-Monte Carlo Software

### 2022-10-31 MCQMC2022QMCSoftwareArticle.tex
Office Computer Updates

%
% sample root file for your "contribution" to a contributed volume
%
% Use this file as a template for your own input.
%
%%%%%%%%%%%%%%%% Springer %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[graybox,footinfo]{svmult}

@@

\usepackage[colorlinks=true,linkcolor=black,citecolor=black,urlcolor=blue]{hyperref}
\urlstyle{same}
\usepackage{bookmark}
\pdfstringdefDisableCommands{\def\and{, }}
\makeatletter % to avoid hyperref warnings:
\providecommand*{\toclevel@author}{999}
\providecommand*{\toclevel@title}{0}
\makeatother

@@

\newcommand{\QMCPYnorm}[2][{}]{\ensuremath{\left \lVert #2 \right \rVert}_{#1}}
\newcommand{\QMCPYnormnorm}[2][{}]{\ensuremath{\lVert #2 \rVert}_{#1}}
\newcommand{\QMCPYbignorm}[2][{}]{\ensuremath{\bigl \lVert #2 \bigr \rVert}_{#1}}
\newcommand{\QMCPYBignorm}[2][{}]{\ensuremath{\Bigl \lVert #2 \Bigr \rVert}_{#1}}
\newcommand{\QMCPYabs}[1]{\ensuremath{{\left \lvert #1 \right \rvert}}}
\newcommand{\QMCPYbigabs}[1]{\ensuremath{{\bigl \lvert #1 \bigr \rvert}}}
\newcommand{\QMCPYBigabs}[1]{\ensuremath{{\Bigl \lvert #1 \Bigr \rvert}}}
\newcommand{\QMCPYbiggabs}[1]{\ensuremath{{\biggl \lvert #1 \biggr \rvert}}}

@@

\definecolor{orange}{rgb}{1.0,0.3,0.0}
\definecolor{violet}{rgb}{0.75,0,1}
\newcommand{\frednote}[1]{  {\textcolor{red}  {\mbox{**Fred:} #1}}}
\newcommand{\scnote}[1]{ {\textcolor{blue}  {\mbox{**SC:} #1}}}
\newcommand{\tonynote}[1]{ {\textcolor{orange}  {\mbox{**Tony:} #1}}}

@@

\newcommand{\AGSComment}[1]{{\color{cyan} Aleksei: #1}}
\newcommand{\FJHComment}[1]{{\color{magenta} Fred: #1}}
\newcommand{\JRComment}[1]{{\color{violet} Jag: #1}}

\newcommand{\hmu}{\widehat{\mu}}
\newcommand{\IID}{\textup{IID}}
\newcommand{\LD}{\textup{LD}}
\newcommand{\IMP}{\textup{IS}}
\newcommand{\unif}{\textup{unif}}
\newcommand{\IIDsim}{\overset{\IID}{\sim}}
\newcommand{\LDsim}{\overset{\LD}{\sim}}
\newcommand{\Msim}{\overset{\textup{M}}{\sim}}
\newcommand{\dig}{\textup{dig}}
\newcommand{\mA}{\mathsf{A}}
\newcommand{\mSigma}{\mathsf{\Sigma}}
\newcommand{\mLambda}{\mathsf{\Lambda}}
\newcommand{\mI}{\mathsf{I}}
\newcommand{\cube}{[0,1]^d}
\newcommand{\Kum}{\textup{Kum}}

\iffalse
Here are some of the main suggestions, other grammatical / wording suggestions are in the reports

* General ~ Sou-Cheng DONE
Expand section 2 and talk more about how QMCPy combines the best of available QMC softwares
Fix LatNet Builder reference Fred
Add tensorflow

* Discrete Distribution ~ Fred
Discuss difference between stratification and RQMC
Comment on the name DiscreteDistribution, the referee and Dirk Nuyens found this confusing Fred done
List criterion used to construct Lattice sequences Aleksei done
Clarify that the better uniformity is for embedded Lattices and is not true in general Fred done

* Code / Choice of Parameters ~ Aleksei, Jags check
Show how the user can customize Sobol' and Lattice generative vectors
Make parameters more explicit e.g. Sobol(2) --> Sobol(dimension=2), and explain more about what the code does for non-python / QMCPy users

* Integrand / Stopping Criterion ~ Jags done, Sou-Cheng check
Explain why we use the trapezoidal rule for AsianOption and how we chose the drift for the importance sampling example? 
Better explanation of CubQMCSobolG algorithm

* Importance Sampling ~ Aleksei done, Fred check
Provide better discussion of importance sampling and justification for the example
Motivation for successive change of variables
Explain the Kumaraswamy distribution

\fi

@@

\title*{Quasi-Monte Carlo Software}
\authorrunning{S.-C.\ T.\ Choi et al.}
\author{Sou-Cheng T. Choi \and Fred J. Hickernell \and R. Jagadeeswaran \and Michael J. McCourt \and Aleksei G. Sorokin}
\institute{Sou-Cheng T. Choi \at Department of Applied Mathematics, Illinois Institute of Technology,\\ RE 220, 10 W.\ 32$^{\text{nd}}$ St., Chicago, IL 60616; and Kamakura Corporation, 2222 Kalakaua Ave, Suite 1400, Honolulu, HI 96815 \email{schoi32@iit.edu}
\and
Fred J. Hickernell \at Center for Interdisciplinary Scientific Computation and \\
Department of Applied Mathematics, Illinois Institute of Technology \\ RE 220, 10 W.\ 32$^{\text{nd}}$ St., Chicago, IL 60616 \email{hickernell@iit.edu}
\and
R. Jagadeeswaran \at
Department of Applied Mathematics, Illinois Institute of Technology,\\ RE 220, 10 W.\ 32$^{\text{nd}}$ St., Chicago, IL 60616 \email{jrathin1@iit.edu}; and \\Wi-Tronix LLC, 631 E Boughton Rd, Suite 240, Bolingbrook, IL 60440
\and
Michael J. McCourt \at SigOpt, an Intel company, \\
100 Bush St., Suite 1100, San Francisco, CA 94104
\email{mccourt@sigopt.com}
\and 
Aleksei G. Sorokin \at
Department of Applied Mathematics, Illinois Institute of Technology,\\ RE 220, 10 W.\ 32$^{\text{nd}}$ St., Chicago, IL 60616 \email{asorokin@hawk.iit.edu}}

@@

\abstract{Practitioners wishing to experience the efficiency gains from using low discrepancy sequences need correct, robust, well-written software.  This article, based on our MCQMC 2020 tutorial, describes some of the better quasi-Monte Carlo (QMC) software available.  We highlight the key software components required  by QMC to approximate multivariate integrals or expectations of functions of vector random variables.  We have combined these components in QMCPy, a Python open-source library, which we hope will draw the support of the QMC community.  Here we introduce QMCPy.}

\section{Introduction} \label{sec:intro}

Quasi-Monte Carlo (QMC) methods promise great efficiency gains over independent and identically distributed (IID) Monte Carlo (MC) methods.  In some cases, QMC  achieves one hundredth of the error of IID MC in the same amount of time (see Figure \ref{fig:sc_comp}). Often, these efficiency gains are obtained simply by replacing IID sampling with low discrepancy (LD) sampling, which is the heart of QMC.

Practitioners might wish to test whether QMC would speed up their computation; doing so would be made easier with access the best QMC algorithms available.  Theoreticians or algorithm developers might want to demonstrate their ideas on various use cases to show their practical value.

This tutorial points to some of the best QMC software available.  Then we describe  QMCPy \cite{QMCPy2020a}\footnote{QMCPy is in active development. This article is based on version 1.0.1 on PyPI.}, which is crafted to be a community-owned Python library that combines the best QMC algorithms and interesting use cases from various authors under a common user interface.

The model problem for QMC is approximating a multivariate integral,
\begin{equation} \label{eq:integral}
	\mu := \int_\calT g(\bst) \, \lambda(\bst) \, \D \bst,
\end{equation}
where $g$ is the integrand, and $\lambda$ is a non-negative weight.  If $\lambda$ is a probability distribution (PDF) for the random variable $\bsT$, then $\mu$ is the mean of $g(\bsT)$.  Regardless, we perform a suitable variable transformation to interpret this integral as the  mean of a function of a multivariate, standard uniform random variable:
\begin{equation} \label{eq:fintegral}
	\mu = \bbE[f(\bsX)] =  \int_{\cube}  f(\bsx) \,  \D \bsx , \qquad \bsX \sim \calU\cube.
\end{equation}

QMC approximates the population mean, $\mu$,  by a sample mean,
\begin{equation} \label{eq:samplemean}
	\hmu := \frac 1n \sum_{i=0}^{n-1} f(\bsX_i), \qquad \bsX_0, \bsX_1, \ldots \Msim \calU\cube.
\end{equation}
The choice of the sequence $\{\bsX_i\}_{i=0}^\infty$ and the choice of $n$ to satisfy  the prescribed error requirement,
\begin{equation} \label{eq:err_req}
	\QMCPYabs{\mu - \hmu} \le \varepsilon \qquad \text{absolutely or with high probability},
\end{equation} 
are important decisions, which  QMC software helps the user make.

Here, the notation $\Msim$ means that the sequence mimics the specified, target distribution, but not necessarily in a probabilistic way.  We  use this notation in two forms:  $\IIDsim$ and $\LDsim$.

IID sequences must be random. The position of any point is not influenced by any other, so clusters and gaps occur.  A randomly chosen subsequence of an IID sequence is also IID.  When we say that $\bsX_0, \bsX_1, \ldots \IIDsim F$ for some distribution $F$, we mean that for any positive integer $n$, the  multivariate probability distribution of $\bsX_0, \ldots, \bsX_{n-1}$ is the product of the marginals, specifically,
\begin{equation*}
	F_{n}(\bsx_0, \ldots, \bsx_{n-1}) = F(\bsx_0) \cdots  F(\bsx_{n-1}).
\end{equation*}
When IID points are used to approximate $\mu$ by the sample mean, the root mean squared error is $\calO(n^{-1/2})$.  Figure \ref{fig:comparePts} displays IID uniform points, $\bsX^{\IID}_0, \bsX^{\IID}_1, \ldots \IIDsim \calU[0,1]^2$, i.e.,  the target distribution is $F_\unif(\bsx) = x_1 x_2$.

\begin{figure}[t]
	\includegraphics[height=5.8cm]{QMCSoftwareArticle/vfigs/dd_iid_uniform_pts.eps}
	\quad
	\includegraphics[height=5.8cm]{QMCSoftwareArticle/vfigs/dd_sobol_pts.eps}
	\caption{IID points (left) contrasted with LD points (right).  The LD points cover the square more evenly.}
	\label{fig:comparePts}
\end{figure}

LD sequences may be deterministic or random, but each point is carefully coordinated with the others so that they fill the domain well.  Subsequences of LD sequences are generally not LD.  When we say that $\bsX_0, \bsX_1, \ldots \LDsim \calU\cube$, we mean that for any positive integer $n$,  the \emph{empirical distribution} of $\bsX_0, \ldots, \bsX_{n-1}$, denoted $F_{\{\bsX_i\}_{i=0}^{n-1}}$,  approximates the uniform distribution, $F_\unif$, well (relative to $n$).  (The empirical distribution of a set assigns equal probability to each point.)

A measure of the difference between the empirical distribution of a set of points and the uniform distribution is called a \emph{discrepancy} and is denoted $D(\{\bsX_i\}_{i=0}^{n-1})$ \cite{DicEtal14a,Hic97a,Hic99a,Nie92}.  This is the origin of the term ``low discrepancy'' points or sequences.  LD points by definition have a smaller discrepancy than IID points.  Figure \ref{fig:comparePts} contrasts IID uniform points with LD points, $\bsX^{\LD}_0, \bsX^{\LD}_1 \ldots \LDsim \calU[0,1]^2$, in this case, linearly scrambled and digitally shifted Sobol' points.

The error in using the sample mean to approximate the integral can be bounded according to the Koksma-Hlawka inequality and its extensions \cite{DicEtal14a,Hic97a,Hic99a,Nie92} as the product of the discrepancy of the sampling sequence and the variation of the integrand, denoted $V(\cdot)$:
\begin{equation}
	\QMCPYabs{\mu - \hmu} = \QMCPYbiggabs{\int_{\cube} f(\bsx) \, \D \bigl(F_\unif - F_{\{\bsX_i\}_{i=0}^{n-1}}\bigr) (\bsx)} \le D\bigl(\{\bsX_i\}_{i=0}^{n-1}\bigr) V(f).
\end{equation} 
The variation is a (semi-) norm of the integrand in a suitable Banach space.  The discrepancy corresponds to the norm of the error functional for that Banach space.  For typical Banach spaces, the discrepancy of LD points is $\calO(n^{-1+\epsilon})$, which is a higher convergence order than for IID points.  For details, the reader is pointed to the references.

Here, we expect the reader to see in Figure \ref{fig:comparePts} that the LD points cover the integration domain more evenly than IID points.  LD sampling can be thought of as a more even distribution of the sampling sites than IID.  LD sampling is similar to stratified sampling.  In the examples below, the reader will see the demonstrably smaller cubature errors arising from using LD points.

In the sections that follow, we first overview available QMC software.  We next describe an architecture for good QMC software, i.e., what the key components are and how they should interact.  We then describe how we have implemented this architecture in QMCPy.  Finally, we summarize further directions that we hope QMCPy and other QMC software projects will take.  Those interested in following or contributing to the development of QMCPy are urged to visit the GitHub repository at \href{https://github.com/QMCSoftware/QMCSoftware}{\nolinkurl{https://github.com/QMCSoftware/QMCSoftware}}.

We have endeavored to be as accurate as possible at the time of writing this article.  We hope that progress in QMC software development will make this article obsolete in a good way in the coming years.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Available Software for QMC} \label{sec:available} 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
QMC software spans  LD sequence generators, cubatures, and applications.  We review the better-known software collections, recognizing that some software overlaps multiple categories. Whenever applicable, we state each library's accessibility in QMCPy, or contrast its functionalities with QMCPy's --- where we lag behind in QMCPy, we strive to catch up in the near future.

Software focusing on generating high-quality LD sequences and their generators includes the following, listed in alphabetical order:

@@

% https://tex.stackexchange.com/questions/74279/how-to-add-bullets-to-description-lists
	\item[BRODA] Commerical and non-commercial software developed jointly with I.M.~Sobol' in C++, MATLAB, and Excel~\cite{BRODA20a}. BRODA can generate Sobol' sequences up to 65,536 dimensions. In comparison, QMCPy supports Sobol' sequences up to 21,201 dimensions.	
	\item[Burkhardt] Various QMC software for generating van der Corput, Faure, Halton, Hammersley,  Niederreiter, or Sobol' sequences in C, C++, Fortran, MATLAB, or Python~\cite{Bur20a}. In QMCPy, we have implemented digital net, lattice, and Halton generators.
	\item[LatNet Builder] The successor to \emph{Lattice Builder}~\cite{LEcMun14a}, LatNet Builder is a C++ library with Python and Java interfaces (in SSJ below) for generating vectors/matrices for lattices and digital nets~\cite{LatNet,LEcEtal22a}. QMCPy contains a module to parse the generators from LatNet Builder for compatibility with our LD point generators. %contains a \href{https://qmcpy.readthedocs.io/en/latest/demo_rst/latnetbuilder.html}{demo for LatNet Builder}. 
	\item[MATLAB] Commercial software \cite{MAT9.10}, which contains Sobol' and Halton sequences in the Statistics and Machine Learning Toolbox. Both generators can be applied in conjunction with MATLAB's Parallel Computing Toolbox to accelerate their execution speed. The maximum dimension of the Sobol' sequences is restricted to 1,111, which is relatively small, yet sufficient for most applications. 
	\item[MPS] Magic Point Shop contains lattices and Sobol' sequences in Python, Matlab, and C++ \cite{Nuy17a}. We have used MPS as a starting point for developing LD generators in QMCPy.
	\item[Owen] Randomized Halton sequences with dimensions up to 1,000 \cite{Owe20a} and scrambled Sobol' sequences with dimensions up to 21,021  \cite{owen1998scrambling, owen2021r}  in R. QMCPy currently supports Owen's Halton randomization method, and we plan to implement Owen's nested uniform scrambling for digital nets in the near future.
	\item[PyTorch] Open-source Python library for deep learning, with scrambled or unscrambled Sobol' sequences \cite{NEURIPS2019_9015,PyTorch}. PyTorch enables seamless utilization of Graphics Processing Units (GPUs) or Field Programmable Gate Arrays (FPGAs).
	\item[QMC.jl] LD Sequences in Julia \cite{Rob20a}. Julia~\cite{bezanson2012julia} is an interpreted language similar to Python and R in terms of ease of use, but is designed to run much faster.
	\item [qrng] Randomized Sobol', Halton, and Korobov sequences in R~\cite{QRNG2020}. The default Halton randomization in QMCPy utilizes the methods from qrng. 
	%translated qrng to Python; see  \url{https://github.com/QMCSoftware/qrngpy}. The Halton randomization from qrng is also optionally available in QMCPy. \AGSComment{We don't really use qrng besides the QRNG Halton randomization.} In addition, the C source files in qrng can be compiled and accessed in QMCPy as one of the backends. In some cases, our version of qrng with its compiled code in QMCPy is more efficient than the original software in R \scnote{Why or how? By taking advantage of MRG63k3a?}; see the demo, \href{https://qmcpy.readthedocs.io/en/latest/demo_rst/quasirandom_generators.html}{\emph{Quasi-Random Sequence Generator Comparison}}, in the QMCPy documentation~\cite{QMCPyDocs}. 
	\item[SciPy] Scientific computing library in Python with Latin hypercube,  Halton, and Sobol' generators \cite{2020SciPy-NMeth}.
	\item[TF Quant Finance] Google's Tensorflow  deep-learning library~\cite{tensorflow2015-whitepaper} specialized for financial modeling \cite{tfqf21}. The software contains lattice and Sobol' generators alongside algorithms accelerated useing GPUs, FGPAs, or Tensor Processing Units (TPUs).% as opposed to the Central Processing Units (CPUs).		
\end{description}
Software focusing on QMC cubatures and applications includes the following:
	\item[GAIL] Guaranteed Automatic Integration Library  contains automatic (Q)MC stopping criteria in MATLAB \cite{ChoEtal21a,HCJJ14}. These are iterative procedures for one- or high-dimensional integration that take a user's input error tolerance(s) and determine the number of (Q)MC sampling points necessary to achieve the desired accuracy (almost surely).  Most of GAIL's (Q)MC functions, some with enhancements, are implemented in Python in QMCPy.
	\item[ML(Q)MC] Multi-Level (Q)MC routines in C, C++, MATLAB, Python, and R~\cite{GilesSoft}. We have ported ML(Q)MC functions to QMCPy.
	\item[MultilevelEstimators.jl] ML(Q)MC methods in Julia \cite{Rob21}. The author, Pieterjan Robbe, has contributed cubature algorithms and use cases to QMCPy.
	\item[OpenTURNS] Open source initiative for the Treatment of Uncertainties, Risks'N Statistics \cite{OpenTURNS} written in C++ and Python, leveraging R statistical packages, as well as LAPACK and BLAS for numerical linear algebra.
	\item[QMC4PDE] QMC for elliptic PDEs with random diffusion coefficients in Python \cite{KuoNuy16a}.
	\item[SSJ] Stochastic Simulation with the \texttt{hups} package in Java \cite{SSJ}.
	\item[UQLab] Framework for Uncertainty Quantification in MATLAB \cite{UQLab2014}.  The core of UQLab is closed source, but a large portion of the library is open source. Recently,  UQ[py]Lab, the beta release of UQLab with Python bindings, is  available as Software as a Service (SaaS) via UQCloud~\cite{lataniotis2021uncertainty}.

\end{description}

The sections that follow describe QMCPy \cite{QMCPy2020a}, which is our attempt to establish a framework for QMC software and to combine the best of the above software under a common user interface written in Python 3.  The choice of language was determined by the desire to make QMC software accessible to a broad audience, especially the technology industry.

\section{Components of QMC Software}
QMC cubature can be summarized as follows.  We want to approximate the expectation, $\mu$, well by the sample mean, $\hmu$, where \eqref{eq:integral}, \eqref{eq:fintegral}, and \eqref{eq:samplemean} combine to give
\begin{multline} \label{eq:cubSummary}
	\mu : = \int_\calT g(\bst) \, \lambda(\bst) \, \D \bst  = \bbE[f(\bsX)] = \int_{\cube} f(\bsx) \, \D \bsx \approx \frac 1n \sum_{i=0}^{n-1} f(\bsX_i) =: \hmu, \\
	 \bsX \sim \calU\cube, \ \bsX_0, \bsX_1, \ldots \Msim \calU\cube.
\end{multline}
Moreover, we want to satisfy the error requirement in \eqref{eq:err_req}.
This requires four components, which we implement as QMCPy classes.

	\item[Discrete Distribution]  produces the sequence $\bsX_0, \bsX_1, \dots$ that mimics $\calU[0,1]^d$;

	\item[True Measure] $\bst \mapsto \lambda (\bst) \D \bst$  defines the original measure, e.g., Gaussian or Lebesgue;

	\item[Integrand] $g$  defines the original integrand, and $f$ defines the transformed version to fit the \texttt{DiscreteDistribution}; and

>>>>

% MCQMC 2022 article on the Challenges of Great MCQMC Software
%
% sample root file for your "contribution" to a contributed volume
%
% Use this file as a template for your own input.
%
%%%%%%%%%%%%%%%% Springer %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[graybox,footinfo]{svmult}

@@

\usepackage[colorlinks=true,linkcolor=black,citecolor=black,urlcolor=black]{hyperref}
\urlstyle{same}
\usepackage{bookmark}
\pdfstringdefDisableCommands{\def\and{, }}
\makeatletter % to avoid hyperref warnings:
\providecommand*{\toclevel@author}{999}
\providecommand*{\toclevel@title}{0}
\makeatother

@@

\newcommand{\QMCPYnorm}[2][{}]{\ensuremath{\left \lVert #2 \right \rVert}_{#1}}
\newcommand{\QMCPYnormnorm}[2][{}]{\ensuremath{\lVert #2 \rVert}_{#1}}
\newcommand{\QMCPYbignorm}[2][{}]{\ensuremath{\bigl \lVert #2 \bigr \rVert}_{#1}}
\newcommand{\QMCPYBignorm}[2][{}]{\ensuremath{\Bigl \lVert #2 \Bigr \rVert}_{#1}}
\newcommand{\QMCPYabs}[1]{\ensuremath{{\left \lvert #1 \right \rvert}}}
\newcommand{\QMCPYbigabs}[1]{\ensuremath{{\bigl \lvert #1 \bigr \rvert}}}

@@

\definecolor{orange}{rgb}{1.0,0.3,0.0}
\definecolor{violet}{rgb}{0.75,0,1}
\newcommand{\frednote}[1]{  {\textcolor{red}  {\mbox{**Fred:} #1}}}
\newcommand{\yuhannote}[1]{ {\textcolor{violet}  {\mbox{**Yuhan:} #1}}}
\newcommand{\tonynote}[1]{ {\textcolor{orange}  {\mbox{**Tony:} #1}}}

@@

\newcommand{\AGSComment}[1]{{\color{cyan} Aleksei: #1}}
\newcommand{\FJHComment}[1]{{\color{magenta} Fred: #1}}

\newcommand{\hmu}{\widehat{\mu}}
\newcommand{\IID}{\textup{IID}}
\newcommand{\LD}{\textup{LD}}
\newcommand{\IMP}{\textup{imp}}
\newcommand{\unif}{\textup{unif}}
\newcommand{\IIDsim}{\overset{\IID}{\sim}}
\newcommand{\LDsim}{\overset{\LD}{\sim}}
\newcommand{\dig}{\textup{dig}}
\newcommand{\mA}{\mathsf{A}}
\newcommand{\mSigma}{\mathsf{\Sigma}}
\newcommand{\mLambda}{\mathsf{\Lambda}}
\newcommand{\mI}{\mathsf{I}}

@@

\title*{Quasi-Monte Carlo Software}
\authorrunning{S.-C.\ T.\ Choi et al}
\author{Sou-Cheng Terrya Choi \and Fred J. Hickernell \and R. Jagadeeswaran \and Michael J. McCourt \and Aleksei G. Sorokin}
\institute{Sou-Cheng Terrya Choi \at Department of Applied Mathematics, Illinois Institute of Technology,\\ RE 220, 10 W.\ 32$^{\text{nd}}$ St., Chicago, IL 60616 \email{schoi32@hawk.iit.edu}
\and
Fred J. Hickernell \at Center for Interdisciplinary Scientific Computation and \\
Department of Applied Mathematics, Illinois Institute of Technology \\ RE 220, 10 W.\ 32$^{\text{nd}}$ St., Chicago, IL 60616 \email{hickernell@iit.edu}
\and
R. Jagadeeswaran \at
\and
Michael J. McCourt \at ???
\and 
Aleksei G. Sorokin \at
Department of Applied Mathematics, Illinois Institute of Technology,\\ RE 220, 10 W.\ 32$^{\text{nd}}$ St., Chicago, IL 60616 \email{asorokin@hawk.iit.edu}}

@@

\abstract{This is the article based on my MCQMC 2020 Tutorial. \AGSComment{Latest code demo at: \href{https://colab.research.google.com/drive/1wijwt0ugsJ9chUHtzSb3lQpc3-N8mCDs?usp=sharing}{https://colab.research.google.com/drive/1wijwt0ugsJ9chUHtzSb3lQpc3-N8mCDs?usp=sharing}}.}

\section{Introduction} \label{sec:intro}
Quasi-Monte Carlo (QMC) methods promise great efficiency gains over independent and identically distributed (IID) Monte Carlo (MC) methods.  In some cases QMC  achieves one hundredth of the error of IID MC in the same amount of time. Often, these efficiency gains are obtained simply by  replacing IID sampling by low discrepancy (LD) sampling, which is the heart of QMC.

As a practitioner, you might wish to test whether QMC would speed your computation.  You would like easy access to the best QMC algorithms available.  As a theoretician or an algorithm developer, you would want to demonstrate your best ideas on  various  use cases to show their practical value.

This tutorial points to some of the best QMC software available.  Moreover, we describe QMCPy \cite{QMCPy2020a}, which is designed to be a community owned Python library that combines the best QMC algorithms from various authors under a common user interface.

The model problem for QMC is approximating an integral,
\begin{equation} \label{eq:integral}
	\mu := \int_\calT g(\bst) \, \lambda(\bst) \, \D \bst,
\end{equation}
where $g$ is the integrand, and $\lambda$ is a non-negative weight.  We use $\mu$ to denote the value of this integral because we interpret it as the population mean of a random variable after a suitable variable transformation:
\begin{equation} \label{eq:fintegral}
	\mu = \bbE[f(\bsX)] = \int_\calX f(\bsx) \, \varrho(\bsx) \, \D \bsx =  \int_\calX f(\bsx) \,  \D F(\bsx) ,
\end{equation}
where $\varrho$ is a probability density with corresponding probability distribution $F$.  In practice, $\calX$ often corresponds to the unit cube, $[0,1]^d$, and $F$ corresponds to the uniform distribution.

QMC approximates this population mean by a sample mean,
\begin{equation} \label{eq:samplemean}
	\hmu := \frac 1n \sum_{i=0}^{n-1} f(\bsX_i), \qquad \bsX_0, \bsX_1, \ldots \sim F.
\end{equation}
The choice of this sequence and the choice of $n$ to satisfy  the prescribed error tolerance are important decisions, which  QMC software helps the user make.

Here, the notation ``$\sim$'' means that the sequence mimics the specified, target distribution, but not necessarily in a probabilistic way.  We  use this notation in two forms:  $\IIDsim$ and $\LDsim$.

IID sequences must be random. The position of any point is not influenced by any other, so clusters and gaps occur.  A subsequence of IID points chosen randomly is also IID.  When we say that $\bsX_0, \bsX_1, \ldots \IIDsim F$, we mean that for any positive integer $n$, the  multivariate probability distribution of $\bsX_0, \ldots, \bsX_{n-1}$ is the product of the marginals, specifically,
\begin{equation*}
	F_{n}(\bsx_0, \ldots, \bsx_{n-1}) = F(\bsx_0) \cdots  F(\bsx_{n-1}).
\end{equation*}
When IID points are used to approximate $\mu$ by the sample mean, the error is $\calO(n^{-1/2})$.  Figure \ref{fig:comparePts} displays IID uniform points, $\bsX^{\IID}_0, \bsX^{\IID}_1, \ldots \IIDsim \calU[0,1]^2$, where the target distribution is $F(\bsx) = x_1 x_2$.

\begin{figure}
	\includegraphics[height=5cm]{QMCSoftwareArticle/figs/dd_iid_uniform_pts.png}
	\qquad
	\includegraphics[height=5cm]{QMCSoftwareArticle/figs/dd_sobol_pts.png}
	\caption{IID points contrasted with LD points.  The LD points cover the square more evenly.}
	\label{fig:comparePts}
\end{figure}

LD sequences may be deterministic or random, but each point is carefully coordinated with the other so that they fill the domain well.  When we say that $\bsX_0, \bsX_1, \ldots \LDsim F$, we mean that for any positive integer $n$,  the \emph{empirical distribution} of $\bsX_0, \ldots, \bsX_{n-1}$, denoted $F_{\{\bsX_i\}_{i=0}^{n-1}}$,  approximates the target distribution, $F$, well (relative to $n$).  (The empirical distribution of a set assigns equal probability to each point.)  The measure of the difference between the empirical distribution of a set of points and the target distribution is called a \emph{discrepancy} and is denoted $D(\{\bsX_i\}_{i=0}^{n-1}, F)$.  This is the origin of the term ``low discrepancy'' points or sequences.  LD points by definition have a smaller discrepancy than IID points.  Figure \ref{fig:comparePts} contrasts IID uniform points with LD points, $\bsX^{\LD}_0, \bsX^{\LD}_1 \ldots \LDsim \calU[0,1]^2$, in this case linearly scrambled Sobol' points. For nearly all LD sequences, the target distribution is $\calU[0,1]^d$.

The error in using the sample mean to approximate the integral can be bounded according to the Koksma-Hlawka inequality and its extensions as the product of the discrepancy of the sampling sequence and the variation of the integrand, denoted $V(\cdot)$:
\begin{equation}
	\QMCPYabs{\mu - \hmu} = \QMCPYabs{\int_{\calX} f(\bsx) \, \D (F - F_{\{\bsX_i\}_{i=0}^{n-1}}) (\bsx)} \le D(\{\bsX_i\}_{i=0}^{n-1}, F) V(f),
\end{equation} 
The variation is a (semi-) norm of the integrand in a suitable Banach space.  The discrepancy corresponds to the norm of the error functional for that Banach space.  For typical Banach spaces, the discrepancy of LD points is $\calO(n^{-1+\epsilon})$, which is a higher convergence order than for IID points.  For details, the reader is referred to the references.  For our purposes, we expect the reader to see in Figure \ref{fig:comparePts} that the LD points cover the integration domain more evenly than IID points.  In the examples below the reader will see the demonstrably smaller cubature errors arising from using LD points.

In the sections that follow we first overview available QMC software.  We next describe an architecture for good QMC software, i.e., what are the key components and how should they interact.  We then describe how we have implemented this architecture in QMCPy.  Finally, we summarize further directions that we hope QMCPy and related software projects will take.  Those interested in following the development of QMCPy or even contributing to that development are urged to visit the GitHub repository at \href{https://qmcsoftware.github.io/QMCSoftware/}{\nolinkurl{https://qmcsoftware.github.io/QMCSoftware/}}.

\section{Available Software for QMC} \label{sec:available} 
QMC software spans  LD sequence generators, algorithms, and applications.  We review the better known software collections, recognizing that some software overlaps multiple categories.
Software focusing on generating high quality LD sequences  or their generators includes
\begin{description}[format=\textup,format=\textbf]
% https://tex.stackexchange.com/questions/74279/how-to-add-bullets-to-description-lists
	\item[BRODA] Sobol' sequences in C, MATLAB, and Excel \cite{BRODA20a},
	\item[Burkhardt] various QMC software in C++, Fortran, MATLAB, \& Python \cite{Bur20a},
	\item[LatNet Builder] Generating vectors/matrices for lattices and digital nets \cite{LEcEtal22a,LatNet},
	\item[MATLAB] Sobol' and Halton sequences, commercial \cite{MAT9.9},
	\item[MPS] Magic Point Shop, lattices and Sobol' sequences \cite{Nuy17a},
	\item[Owen] Randomized Halton sequences in R \cite{Owe20a},
	\item[PyTorch] Scrambled Sobol' sequences \cite{PyTorch},
	\item[QMC.jl] LD Sequences in Julia \cite{Rob20a}, and
	\item [qrng]  Sobol', Halton, and Korobov sequences in R \cite{QRNG2020}.
\end{description}
Software focusing on QMC algorithms and applications includes
\begin{description}[format=\textup,format=\textbf]
	\item[GAIL] Automatic (Q)MC stopping criteria in MATLAB \cite{ChoEtal20a},
	\item[ML(Q)MC] Multi-Level (Quasi-)Monte Carlo routines in C++, MATLAB, Python, and R \cite{GilesSoft},
	\item[OpenTURNS] Open source initiative for the Treatment of Uncertainties, Risks 'N Statistics in Python \cite{OpenTURNS},
	\item[QMC4PDE] QMC for elliptic PDEs with random diffusion coefficients \cite{KuoNuy16a},
	\item[SSJ] Stochastic Simulation in Java \cite{SSJ}, and
	\item[UQLab] Framework for Uncertainty Quantification in MATLAB \cite{UQLab2014}.
\end{description}

The sections that follow describe QMCPy \cite{QMCPy2020a}, which is our attempt to combine the best of the above software under a common user interface written in Python 3.  The choice of language was determined by the desire to make QMC software accessible to a broad audience, especially the tech industry.

\section{Components of QMC Software}
QMC cubature can be summarized as follows.  We want to approximate the integral oe expectation, $\mu$, well by the sample mean, $\hmu$, where \eqref{eq:integral}, \eqref{eq:fintegral}, and \eqref{eq:samplemean} combine to give
\begin{multline} \label{eq:cubSummary}
	\mu : = \int_\calT g(\bst) \, \lambda(\bst) \, \D \bst  = \bbE[f(\bsX)] = \int_\calX f(\bsx) \, \varrho(\bsx) \, \D \bsx \approx \frac 1n \sum_{i=0}^{n-1} f(\bsX_i) =: \hmu, \\
	 \bsX_0, \bsX_1, \ldots \sim F.
\end{multline}
This requires four components, which we implement as QMCPy classes.

@@

	\item[Discrete Distribution] that produces $\bsX_0, \bsX_1, \dots$ mimicking the distribution $F$, which typically is $\calU[0,1]^d$;

	\item[True Measure] $\bst \mapsto \lambda (\bst) \D \bst$ that defines the original integral, e.g., Gaussian or Lebesgue;

	\item[Integrand] $g$ that  defines the original integral, plus the transformed version, $f$, to fit the discrete distribution; and

	\item[Stopping Criterion] that determines how large $n$ should be to ensure that $\QMCPYabs{\mu - \hmu_n} \le \varepsilon$.
\end{description}

The software libraries referenced in Section \ref{sec:available} provide one or more of these components. QMCPy combines multiple examples of them all under and object oriented framework. Each example is implemented as a concrete class that realizes the properties and methods required by the abstract class for that component. The following sections detail descriptions and use cases for each component and select implementations.

Thorough documentation of all QMCPy classes can be found in \cite{QMCPyDocs}. Demonstrations of how QMCPy work are given in Google Colab notebooks \cite{QMCPyTutColab2020,QMCPyTutColab2020_paper} \AGSComment{Update \cite{QMCPyTutColab2020_paper}'s link}. The project may be installed from PyPI into your Python 3 environment via the command \texttt{pip install qmcpy}. In the codes that follows, we assume QMCPy has been imported alongside Numpy \cite{numpy} 
%and Scipy \cite{scipy} 
via
\lstinputlisting[style=Python]{QMCSoftwareArticle/python/import.txt}
